---
title: "First Course in R"
author: "Jeremy Ash"
date: "April 9, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

I am a fourth year Bioinformatics student in Dr. Fourches's lab.  I developed a R package
called [chemmodlab](https://cran.r-project.org/web/packages/chemmodlab/index.html).

Some of these materials have been based on Brian Reich's ST540 course materials available
in the Introduction to R section [here](https://www4.stat.ncsu.edu/~reich/ABA/code.html).

# R Basics

## Basic arithmetic

You can do any basic arithmetic in R.  It can be used as your personal calculator! Try submitting a few commands.

```{r}
(3-1)*4
2^4
```

## Variables

Use the arrow operator to assign a value to a variable.
All of these variables are stored as objects in your R Environment.  

```{r}
x <- 2
```

You can see all of the objects
that live in your environment.

```{r}
ls()
```


## Data structures

### Vectors

The most basic data structure in R is a vector.  Numeric vectors can be constructed
in a number of ways.

```{r}
x1 <- c(1,3,5,7)
x1

x2 <- 1:5
x2
```

### Functions

Base R comes with a number of "functions" that will perform operations on objects in your environment.
In most cases, these are just commands that are wrappers for more R code.  Same of the base R functions
are written in C.

```{r}
# Take a random sample of size 2 from x2
sample(x2, 2)
sample
```

Anytime you do not know how to use a function that you find on the internet, you can almost always
find detailed documentation in R.

```{r, eval=FALSE}
?sum
# The ?? command will search for all functions that match a string
??colMeans
```

### More numeric vectors

Now a few other ways to create numeric vectors using functions.....

```{r}
x3 <- seq(10,15,0.5)
x3

x4 <- rep(5,5)
x4
```

### Other vector types

There are three main types of vectors: *numeric*, *character*, and *factor*.

```{r}
# numeric vectors
class(x1)
x1

# character vectors
x5 <- c("foo", "bar")
class(x5)
x5

# factor vectors
x6 <- as.factor(c("high", "mid", "low"))
class(x6)
x6
as.numeric(x6)
```

### Arithmetic operations on numeric vectors

Arithmetic operations occur on numeric vectors elementwise.

```{r}
x2*3
x2^2
```

Vectors and be combined through arithmetic operation, if they are the same length.

```{r}
length(x1)
length(x2)
x12 <- x1+x2

# There was only a warning
# The sum of the first four elements of both vectors was returned
# See https://cran.r-project.org/doc/manuals/r-release/R-intro.html#The-recycling-rule for more details
x12
```

```{r}
length(x2)
length(x4)
x2+x4
```

### Matrices

A matrix is a two-dimensional object with rows and columns. Data are often stored as matrices where rows are observations and columns are variables.

Vectors can be combined to make a matrix.

```{r}
y <- cbind(x2,x4)
y
dim(y)
```

You can also fill a matrix with values in major column order.

```{r}
mat <- matrix(1:6,3,2)
mat
```

# Cheminformatics example

## Load the Comma separated variable (CSV) data set

```{r}
df <- read.csv("C:/Users/jrash/ubuntu_share/Google Drive/FourchesLab/CH795/ERK2_3D_des_and_pki.csv")
class(df)
```

The CSV was read as a data frame.  A data frame is just like a matrix, except it can 
store multiple vector types in its rows/columns.

This data frame is way too big to look at it in the R environment.

```{r}
dim(df)
```

It is often helpful to just
take a peak at the object to make sure it is structured the way you expect it to be.

```{r}
head(df)
```

The dollar sign is used to extract vectors from the data frames columns.

```{r}
head(df$Set)
class(df$Set)
head(df$pki)
class(df$pki)
```

## Histograms

Make a histogram of each descriptor and pki.

```{r}
hist(df$Cst.Sigma)
hist(df$VdWVolumes.WV)
hist(df$pki)
```

## Summary measures

Compute some summary measures of the Cst.Sigma descriptor.

```{r}
mean(df$Cst.Sigma)
sd(df$Cst.Sigma)
range(df$Cst.Sigma)
```

Say we wanted to compute a new descriptor, the average of Cst.Sigma and VdWVolumes.WV.

```{r}
head((df$Cst.Sigma + df$VdWVolumes.WV)/2)
```

## Plot the data

```{r}
plot(df$Cst.Sigma[1:500], df$VdWVolumes.WV[1:500])
```

We can modify a plot by passing options, such as the axis labels.

```{r}
plot(df$Cst.Sigma[1:500], df$VdWVolumes.WV[1:500],
     xlab="SD of distance from centroid",ylab="Van Der Wall volumes WV",
     main="Relationship between descriptors")
```

Now lets color ligand conformations by pki.

```{r}
colfunc <- colorRampPalette(c("red", "green"))
pki.col <- colfunc(2)[as.numeric(cut(df$pki,breaks = 2))]
plot(df$Cst.Sigma[1:500], df$VdWVolumes.WV[1:500],
     xlab="SD of distance from centroid",ylab="Van Der Wall volumes WV",
     main="Relationship between descriptors and pki",
     col = pki.col)
```

Lets plot the whole dataset!

```{r}
pki.col <- colfunc(10)[as.numeric(cut(df$pki, breaks = 10))]
plot(df$Cst.Sigma, df$VdWVolumes.WV,
     xlab="SD of distance from centroid",ylab="Van Der Wall volumes WV",
     main="Relationship between descriptors and pki",
     col = pki.col)
```

There is a lot of overplotting going on here that is making the pattern in the data unclear.  Lets try reducing the point size.

```{r}
pki.col <- colfunc(10)[as.numeric(cut(df$pki, breaks = 10))]
plot(df$Cst.Sigma, df$VdWVolumes.WV,
     xlab="SD of distance from centroid",ylab="Van Der Wall volumes WV",
     main="Relationship between descriptors and pki",
     col = pki.col, cex = .2)
```

And changing the plotted points from circles to points.

```{r}
pki.col <- colfunc(10)[as.numeric(cut(df$pki, breaks = 10))]
plot(df$Cst.Sigma, df$VdWVolumes.WV,
     xlab="SD of distance from centroid",ylab="Van Der Wall volumes WV",
     main="Relationship between descriptors and pki",
     col = pki.col, cex = .2, pch = 20)
```


You can make really pretty graphics using ggplot2.   This is a library of functions that doesn't
come installed in base R.  So you will need to install the library from the CRAN server and load
into your workspace.

```{r, eval=F}
install.packages("ggplot2")
```


```{r}
library(ggplot2)

g <- ggplot(df, aes(x = Cst.Sigma, y = VdWVolumes.WV), alpha=.001)
g <- g + geom_point(aes(colour=pki), size=.5)
g <- g + scale_color_continuous(name = "pki",high='darkgreen', low='red')
g <- g + theme(title = element_text(size = rel(1.5)), 
               axis.text=element_text(size = rel(1.5)),
               axis.title=element_text(size=rel(.8)),
               legend.title=element_text(size = rel(.75)),
               legend.text=element_text(size = rel(1.25)))

print(g)
```


